Prasanna Velisetti
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

library(ggplot2)
library(knitr)
library(dplyr)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
data_wine<-read.csv('wineQualityWhites.csv')
data_wine<-transform(data_wine, quality = factor(quality, 
    levels = c("Very Low","Low","Normal","High","Very High"), ordered = TRUE))
data_wine<-data_wine[,2:13]
data_wine_scaled<-scale(data_wine[,1:11])
```

> **White Wine Quality**: The dataset contains physio-chemical properties of 
4898 different samples of Portuguese "Vinho Verde" wine. The value for quality 
of wine is based on sensory data (median of at least 3 evaluations made by wine
experts).The dataset will be used to assess the important physio-chemical 
properties that determines the quality of wine.There are no missig values
in the dataset.


# Univariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE,Univariate_Plots}
str(data_wine)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(data_wine)
```


As seen from the summary, all fields are nummeric. Let us look into the 
histograms of each of the fields.

```{r  echo=FALSE,  message=FALSE, warning=FALSE}
qplot(x=fixed.acidity,data=data_wine,geom='histogram',binwidth=0.2)
summary(data_wine$fixed.acidity)
```

Fixed acidity distribution appears to be normal with a mean & median around 6.8
(g/dm^3). 

```{r  echo=FALSE,  message=FALSE, warning=FALSE}
qplot(x=volatile.acidity,data=data_wine,geom='histogram',
      binwidth=0.05)
qplot(x=scale(volatile.acidity),data=data_wine,geom=
        'histogram',binwidth=0.5)+geom_vline(xintercept=3.5,colour='red')

summary(data_wine$volatile.acidity)
```

It can be seen that the distribution slightly right skewed. I wonder what 
happens if we combine both fixed acidity and volatile acidity as in terms of
wine tasting, they both may not be distinguished.

Volatile acidity is standardized/scaled in the second graph to see if there are any 
outliers (>3.5)


```{r  echo=FALSE,  message=FALSE, warning=FALSE}
qplot(x=citric.acid,data=data_wine,geom='histogram',binwidth=0.01) # sudden peak at 0.5 
summary(data_wine$citric.acid)
```

There is a sudden peak in the citric acid concentration around 0.5 (g / dm^3). 

```{r  echo=FALSE,  message=FALSE, warning=FALSE}
qplot(x=residual.sugar,data=data_wine,geom='histogram',binwidth=0.3)+
  scale_x_continuous(trans = "log2") #Bimodal; Right Skewed

summary(data_wine$residual.sugar)

ggplot(data_wine, aes( x = 1, y = residual.sugar ) ) +
geom_jitter(alpha = 0.1 ) +scale_y_continuous(trans = "log2")
               
```


Residual sugar distribution is clearly bimodal. One mode occurs 
around 1.5 and another around 12

One important thing to note here is that the residual sugar gives us a picture 
of unfermented sugar left in the wine. The bimodal distribution indicates that
there is decent demand for both sweet wines and dry wines. Some may like to have
a good taste of wine while some may want to have high alcohol. Here my
assumption is that alcohol and sugar content might be realted. I would like to 
see if their relation.




```{r  echo=FALSE,  message=FALSE, warning=FALSE}
qplot(x=chlorides,data=data_wine,geom='histogram',binwidth=0.005)
summary(data_wine$chlorides)
```

Chloride distribution seem to be right skewed with median around 0.043.

```{r  echo=FALSE,  message=FALSE, warning=FALSE}
qplot(x=free.sulfur.dioxide,data=data_wine,geom='histogram',binwidth=5)

ggplot(data_wine, aes( x = 1, y =free.sulfur.dioxide ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) 

summary(data_wine$free.sulfur.dioxide)
```



```{r  echo=FALSE,  message=FALSE, warning=FALSE}
qplot(x=total.sulfur.dioxide,data=data_wine,geom='histogram',binwidth=10)
ggplot(data_wine, aes( x = 1, y =total.sulfur.dioxide ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) 
summary(data_wine$total.sulfur.dioxide)
```

Both total & free sulfur dioxide distributions are slight right skewed. 
There are some outliers as seen from the box plots. Total sulfur dioxide 
includes free sulfur dioxide as well. Would like to see the relation of 
bound sulfur dioxide which is the difference between total and free sulfur
dioxide and the quality of wines along with the free and total sulfur dioxide.


```{r  echo=FALSE,  message=FALSE, warning=FALSE}
qplot(x=density,data=data_wine,geom='histogram',binwidth=0.001)
summary(data_wine$density)
```

Density ranges from 0.9871 (g / cm^3) to 1.039 (g / cm^3). We see that there 
is not much variation in the density. As we have the different chemical content 
densities given in the dataset, it is most likely that density is highly 
correlated with other features. So I want to explore this further in bi-variate
analysis.

```{r  echo=FALSE,  message=FALSE, warning=FALSE}
qplot(x=pH,data=data_wine,geom='histogram',binwidth=0.05)

ggplot(data_wine, aes( x = 1, y =pH ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) 


summary(data_wine$pH)
```


```{r  echo=FALSE,  message=FALSE, warning=FALSE}
qplot(x=alcohol,data=data_wine,geom='histogram',binwidth=0.5)
summary(data_wine$alcohol)
```

The minimum alcohol content is around 8% and maximum is  around 14%. The 
distribution is right skewed with median = 10.4%



```{r  echo=FALSE,  message=FALSE, warning=FALSE}
qplot(x=quality,data=data_wine) 
summary(data_wine$quality)
```

We see that distribution of wines is highly biased towards normal quality 
wines and there are very rare (5) Very high quality wines and no low quality 
wines. Also around 95% of the data is either Normal or High Quality. It is a 
good idea to remove the Very HIgh and Very Low quality wines from the analysis
to determine which affects the wine quality as the number of datapoints are 
very less.




```{r echo=FALSE,  tidy=FALSE,message=FALSE, warning=FALSE}

data_wine$Bound_sulfur_dioxide=(data_wine$total.sulfur.dioxide 
                                -data_wine$free.sulfur.dioxide)

data_wine$ratio_alcohol_sugar=data_wine$alcohol/data_wine$residual.sugar
data_wine$diff_sugar_citicacid=(data_wine$residual.sugar-
                                  data_wine$citric.acid
                                -data_wine$fixed.acidity)/data_wine$alcohol
data_wine_filtered<-subset(data_wine,quality !='Very High')

```


# Univariate Analysis

### What is the structure of your dataset?
There are 4898 observations with 11 different fields excluding quality field. 
Every field is a continous variable. 

### What is/are the main feature(s) of interest in your dataset?

No judgement has been taken at this stage

### Did you create any new variables from existing variables in the dataset?
1) Bound_sulfur_dioxide which is difference between total sulfur dioxide and 
   the free sulfur dioxide
2) Quality is changed into categorical variable with Very Low (1,2), Low (3,4), 
   Normal (5,6), High (7,8) and Very High (9,10) categories
3) Free sulfur dioxide to Total Sulfur doxide ratio
4) Ratio of alcohol to residual sugar

### Of the features you investigated, were there any unusual distributions? 

1) There is a sudden peak in citric acid concentration aroud 0.5 (g / dm^3)
2) Bimodal distribution for residual sugar

# Bivariate Plots Section


```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots}
M<-cor(data_wine[,-12])
library('corrplot') #package corrplot
corrplot(M, method = "circle") #plot matrix
```


Density is largely correlated (>0.8) to residual.sugar and alcohol 
percentage(<-0.8) 
As density is combined concentrations of its contents, this behaviour 
is expected. Density field can thus be ignored.


```{r  echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data_wine, 
       aes(x=residual.sugar, 
           y=alcohol)) +  geom_point(
             alpha = .2, size = 2,position = 'jitter') +scale_x_continuous(
               limits = c(0, 20))

summary(lm(formula = alcohol ~ residual.sugar, data = 
             subset(data_wine, residual.sugar <=  
                      quantile(data_wine$residual.sugar, 0.999))))

```

As discuseed earlier, there is expected to be relation between Alcohol 
percentage and residual sugar content. It was expected that alcohol and residual
sugar would have a linear relationship given a fixed amount initial sugar 
content but looking at the scatter plot and the
R2 value of the linear regression, the linear relationship is not apparent.This 
makes me think that the intial sugar content might not be the same for the wines
as it depends on the type of grapes. 

The variation of alcohol content decreases with increase in 
residual sugar content. This is because there should be limit on the intial 
sugar content present before fermentation.



```{r echo=FALSE,message=FALSE, warning=FALSE}

qplot(x=quality, y=alcohol, data=data_wine_filtered, geom='boxplot')
qplot(x=quality, y=residual.sugar, data=data_wine_filtered, 
      geom='boxplot')+scale_y_continuous(lim=c(0,40))

```

Alcohol concentration seems to be a good indicator of wine quality as there
are considerable difference in the medians between Normal and High quality wines

Higher and lower quality wines have lower residual sugar content median
compared to that of normal wines. 

It is observed that both Very High and Low quality wines can have lower residual
sugars. Thus Residual sugar might not be good indicator of quality.It would be 
interesting to see if residual sugar content combined with citric acid is a good
feature for quality.


```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x=quality, y =fixed.acidity, data=data_wine_filtered, 
      geom='boxplot')  
qplot(x=quality, y =citric.acid, data=data_wine_filtered, 
      geom='boxplot')+scale_y_continuous(lim=c(0.1,0.5))

qplot(x=quality, y =volatile.acidity, data=data_wine_filtered, 
      geom='boxplot') 

```

Fixed acidity and Citric acid concentration on the other hand seems to be not
much distinguishing the quality of wines.But we can see that the variance of
citric acid decreases with increase in quality.

On the other hand volatile acidity seems to be a good feature.


```{r echo=FALSE,message=FALSE, warning=FALSE}
qplot(x=quality, y =sulphates, data=data_wine_filtered, geom='boxplot')
qplot(x=quality, y =free.sulfur.dioxide, data=data_wine_filtered, 
      geom='boxplot')
qplot(x=quality, y =total.sulfur.dioxide, data=data_wine_filtered, 
      geom='boxplot')
qplot(x=quality, y =free.sulfur.dioxide/total.sulfur.dioxide, 
      data=data_wine_filtered, geom='boxplot')



```


Free sulfur dioxide seems to distinguishing among different 
qualities more than the total sulfur dioxide. We can further analyse to see if
the ratio of free to total sulfur dioxide is a better feature for determining 
quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x=quality, y =chlorides, data=data_wine_filtered, 
      geom='boxplot')+scale_y_continuous(lim=c(0,0.08))


```


There are lot of outliers in Normal quaity wines in terms of chloride 
concentation. If we remove some of the outliers, we can see that with the 
chloride concentration quality decreases. This is expected as the wine starts
to taste salty.



# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation How did the feature(s) of interest vary with other features in the dataset?

1) Alcohol concentration seems to be a good indicator of wine quality as 
there are considerable difference in the medians among Normal, High and 
very high quality wines. With the increase in Alcohol concentration, 
the quality of wine increases

2) As Free sulfur dioxide increases, the quality is seen to be increasing

3) Chlorides seem to be good indicators of wine quality.

4) Residual sugar did not show any trend in the quality


### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

1) Checking the relation between residual sugar and alcohol concentrations.
From the scatter plot, it is observed that higher the residual sugar 
concentration, the narrower the alcohol concentration range becomes. 
With the process of fermentation, natural sugar present in the grapes
is converted to alcohol. Given that there should be a limit on how much 
initial sugar content grapes can contain, depending on how long the 
fermentation process goes, more and more sugar is converted into alcohol. 
So with a lesser residual sugar concentration, it is quite possible to have 
different levels of alcohol but with higher residual sugar content there is 
less possibility to have more alcohol content.

2)Density is strongly related to residual sugar and alcohol content. 
Thus Density can be removed from the list of features as there will be 
redundant information.

### What was the strongest relationship you found?
Density with residual sugar content and alcohol



# Multivariate Plots Section


```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}
qplot(x=quality, y =free.sulfur.dioxide/total.sulfur.dioxide, 
      data=data_wine_filtered, geom='boxplot')
```

Free to total sulfur dioxide ratio should give us an idea on how much 
proportion of added sulfur dioxide is actually avaiable for protection against
microbes. The trend for quality is as expected and more differentiating than
if we take free sulfur dioxide itself.


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data_wine_filtered, aes(x=pH, 
                      y=free.sulfur.dioxide/total.sulfur.dioxide,
                      colour=quality))+  geom_point(shape=1)+geom_point(
                        )+geom_smooth(method = "lm",se = FALSE,
                              size=2)+scale_color_brewer()
```
As the amount of free sulfur dioxide for a given total sulfur dioxide
may depend on the pH of the solution, it is plotted to check if there is any
relation. While there are some trend seen for high quality wins, it is not 
so prominent.


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data_wine_filtered, aes(y=fixed.acidity, 
                      x=residual.sugar,colour=quality)) +  geom_point(
                      ) +geom_point(
                            )+geom_smooth(
                              method = "lm",
                              se = FALSE,
                              size=2)+scale_color_brewer(
                                )+scale_x_continuous(lim=c(0,20))


```

As acidity and sweetness counteract each other, it is interesting to know their
relationship.

We see that with increase in fixed acidity, the residual sugar content 
increases.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?
1)The ratio of free sulfur dioxide to total sulfur dioxide seems to be good
feature that can differentiate well between diffetent qualities of wine.

2)We saw that with increase in fixed acidity, the residual sugar content 
increases. This shows that grapes are picked at a particular time during the
ripening phase as there seems to be an importance to balance out acidity and
residual sugar


### Were there any interesting or surprising interactions between features?
No.


------

# Final Plots and Summary


### Plot One
```{r echo=FALSE,message=FALSE, warning=FALSE, Plot_One}
ggplot(data_wine, 
       aes(x=residual.sugar, 
           y=alcohol)) +  geom_point(
             alpha = .2, size = 2,position = 'jitter') +scale_x_continuous(
               limits = c(0, 20))+labs(
                              x="Residual Sugar (g/dm3)",y
                              ="Alcohol(% by volume)",title='
                              % of Alcohol by Residual Sugar') 
```

### Description One
It is interesting to notice how variation of alcohol percentage decreases with
residual sugar. Even though there is not much linear relationship seen between 
alcohol and residual sugar, the extent to which alcohol percentage can vary
among different wines and the residual sugar content can be related.



### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
qplot(x=quality, y =free.sulfur.dioxide/total.sulfur.dioxide, 
      data=data_wine_filtered, geom='boxplot')+labs(
                              x="Quality",y
                              ="Free sulfur Dioxide/Total sulfur dioxide",
                                title='
                              Ratio of sulfur dioxids by Quality') 
```

### Description Two

Ratio of sulfur dioxides is a good indicator for Quality of wines as it shows
how efficiently the sulfur dioxide is used for protection against microbes

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE,Plot_Three}
qplot(x=quality, y =chlorides,
      data=data_wine_filtered, geom='boxplot')+ scale_y_continuous(
        lim=c(0,0.08))+  labs(
        y="Chlorides",x="Quality",
        title='
                                        Chlorides concentration by Quality')

```

### Description Three

We saw that chlorides concentration is a good feature for 
identifying wine quality.

# Reflection

Most of chemical properties did not correlate with each other. The dataset is 
biased and has very less numner of very high quality wines.A lot of background 
study on different chemical reactions had to be studied to understand the 
realtion between different parameters. While there was background study needed, 
most of relations are easily understandable. It was suprising that there are
both low quality and very high quality wines with high concentration of residual
sugar. 

While different chemical properties of wines are given, there was no information
on the conditions in which the grapes were grown.

There were so many outliers observed in various parts of the analysis and the 
reasons for such anomalies could not be explained based only on the information
given.

More data needs to be collected corresponding to the low and very high quality 
winesto make the dataset unbiased towards different quality of wines
which can then be used for predictive models to predict quality of wine 
given a set of physio-chemical properties.





